---
title: 'Econometrics 2 PS4: Regression Discontinuity'
author: "Carlos T. Estrada Arzamendi"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    extra_dependencies: float
    keep_tex: true
keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(data.table)
library(stargazer)
library(lubridate)
library(modelsummary) # to use datasummary() summary stats
library(haven)
library(MatchIt)
library(dplyr)
library(ggplot2)
#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
options(scipen = 5)
```

# Problem 1: (Sharp) Regression Discontinuity

"Take any dataset with covariate X and outcome Y that are related in
some way. For instance, you can use the data on birth weight and smoking
from here: <http://www.stata.com/texts/eacsap/>, or any other relevant
dataset. Alternatively, feel free to simulate your own data. In any
case, please provide an explanation of your dataset. Construct a placebo
treatment by applying a rule such that Di = 1 when Xi >= x0 for some x0.
That is, modify the outcome variable Yi for those units with Xi >= x0 by
adding a constant treatment effect, for example, add one standard
deviation of the outcome plus some noise (with mean zero). Include an
explanation of what you ended up doing"

```{r data simulation}
# Change to test how RMarkdown works with VSCode
# Change 2--- synced?
# Change 3 --- branch name changed?

# Data Simulation
library(truncnorm)
n = 10000

data = data.table(
  gpa = rtruncnorm(n, a = 0, b = 4, mean = 3, sd = 1),
  fam_income = runif(n, min = 20000, max = 150000)
)
gpa_cutoff = 2

# if gpa is below cutoff, school provides tutoring that increases score
data[, treat := ifelse(gpa < gpa_cutoff, 1, 0)]
data[, outcome_score := gpa*300 + fam_income/1000 + treat*200 + rnorm(n, mean = 0, sd = 50)]

datasummary_skim(data, out = "markdown", title = "Summary Statistics", histogram = F)


```

## 1.1) Plot the outcome by forcing variable (the standard graph showing the discontinuity)


```{r part1}

# Generate input data for output plot

plot1 <- ggplot(data, aes(x = gpa, y = outcome_score, color = as.factor(treat), group = as.factor(treat))) + 
  geom_point() +
  #geom_smooth(aes(fill = as.factor(treat))) +
  geom_smooth(method = "lm",color = "black", formula = y~x) + 
  labs(x = "GPA", y = "Outcome Score") + ggtitle("Outcome by forcing variable") +
  geom_vline(xintercept = gpa_cutoff, linewidth = 1)
plot1

```

## 1.2) Plot the density of the forcing variable

You can also embed plots, for example:

```{r part2, out.width = "75%", fig.align = 'center'}
plot2 <- ggplot(data, aes(gpa)) + 
  geom_histogram(fill = "blue", color = "lightblue", binwidth = 0.05) +
  labs(x = "GPA", y = "Outcome Score") + ggtitle("Density by forcing variable") +
  geom_vline(xintercept = gpa_cutoff, linewidth = 1)
plot2
```

## 1.3) Estimate the effect using a local linear regression


```{r, results='asis', header=FALSE}

reg3 = lm(outcome_score ~ treat + I(gpa-gpa_cutoff) + I(treat*(gpa-gpa_cutoff)) , data = data)
stargazer(reg3, type = "latex", title = "Local Linear Regression", table.placement = "H",
          header=FALSE, no.space = T, omit.stat = "f")

```

## 1.4) Estimate the effect using a local polynomial (of order 2 and 3) regression

```{r order 2, results='asis', header=FALSE}

reg41 = lm(outcome_score ~ treat + I(gpa-gpa_cutoff) + I(treat*(gpa-gpa_cutoff))+
             I((gpa-gpa_cutoff)^2) + I(treat*(gpa-gpa_cutoff)^2), data = data)
#stargazer(reg41, type = "latex", title = "Estimate Effect Using Local Polynomial of Order 2", table.placement = "H", header=FALSE, no.space = T)

```

```{r order 3, results='asis'}
reg42 = lm(outcome_score ~ treat + I(gpa-gpa_cutoff) + I(treat*(gpa-gpa_cutoff))+
             I((gpa-gpa_cutoff)^2) + I(treat*(gpa-gpa_cutoff)^2)+
             I((gpa-gpa_cutoff)^3) + I(treat*(gpa-gpa_cutoff)^3), data = data)
stargazer(reg3, reg41, reg42, type = "latex", title = "Estimate Effect Using Local Polynomial of Order 2 and 3",
          table.placement = "H", header=FALSE, no.space = T, omit.stat = "f")
```
